Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-22 paper.tex
Up nombres y cover letter

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts,'' whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility.

@@

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature {[}campbell\_policy\_2012;
pierson\_when\_1993{]}.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they endorse
meritocratic principles. This distinction aligns with recent findings
indicating that individuals distinguish between how merit is perceived
in society and how it should ideally operate, which in turn shapes their
preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}) Meritocratic beliefs serve as symbolic justifications for unequal
outcomes, particularly when access is stratified by income or social
opportunity. Prior studies have shown that individuals who perceive
higher levels of meritocracy tend to express stronger support for
unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

@@

~

Model 1

Model 2

Model 3

Model 4

Model 5

Model 6

Model 7

Intercept

1.968***

2.015***

2.030***

2.028***

2.169***

1.456***

1.547***

~

(0.021)

(0.035)

(0.035)

(0.035)

(0.070)

(0.097)

(0.115)

Wave (Ref.= 2016)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Wave 2017

-0.167***

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2018

-0.025

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2019

-0.046

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2022

0.285***

~

~

~

~

~

~

~

(0.028)

~

~

~

~

~

~

~~~~~Wave 2023

0.284***

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

Wave

~

-0.121***

-0.125***

-0.128***

-0.128***

-0.128***

-0.129***

~

~

(0.022)

(0.022)

(0.022)

(0.022)

(0.022)

(0.022)

Wave\^{}2

~

0.028***

0.028***

0.029***

0.029***

0.029***

0.029***

~

~

(0.003)

(0.003)

(0.003)

(0.003)

(0.003)

(0.003)

Perception inequality (WE)

~

~

-0.039***

-0.034***

-0.034***

-0.034***

-0.034***

~

~

~

(0.010)

(0.010)

(0.010)

(0.010)

(0.010)

Merit: Effort (WE)

~

~

~

0.070***

0.070***

0.069***

0.069***

~

~

~

~

(0.012)

(0.012)

(0.012)

(0.012)

Merit: Talent (WE)

~

~

~

0.027*

0.027*

0.027*

0.027*

~

~

~

~

(0.012)

(0.012)

(0.012)

(0.012)

Perception inequality (BE)

~

~

~

~

-0.042*

-0.006

-0.033

~

~

~

~

~

(0.018)

(0.018)

(0.018)

Merit: Effort (BE)

~

~

~

~

~

0.186***

0.178***

~

~

~

~

~

~

(0.032)

(0.031)

Merit: Talent (BE)

~

~

~

~

~

0.040

0.019

~

~

~

~

~

~

(0.032)

(0.031)

Controls

No

No

No

No

No

No

Yes

BIC

20780.594

20780.617

20781.113

20726.753

20736.627

20655.172

20749.047

Numb. obs.

8643

8643

8643

8643

8643

8643

8643

Num. groups: individuals

1687

1687

1687

1687

1687

1687

1687

Var: individuals (Intercept)

0.172

0.223

0.219

0.209

0.207

0.182

0.171

Var: Residual

0.530

0.499

0.498

0.495

0.495

0.495

0.495

Var: individuals, wave

~

0.011

0.010

0.009

0.009

0.009

0.009

Cov: individuals (Intercept), wave

~

-0.025

-0.024

-0.021

-0.020

-0.019

-0.019

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05.

@@

~

Model 8

Model 9

Model 10

Model 11

Intercept

1.561***

1.555***

2.187***

2.340***

~

(0.115)

(0.115)

(0.265)

(0.281)

Perception inequality (WE)

-0.035***

-0.035***

-0.035***

-0.035***

~

(0.010)

(0.010)

(0.010)

(0.010)

Merit: Effort (WE)

0.072***

0.074***

0.073***

0.073***

~

(0.013)

(0.012)

(0.012)

(0.012)

Merit: Talent (WE)

0.031*

0.031*

0.030*

0.030*

~

(0.012)

(0.013)

(0.012)

(0.012)

Perception inequality (BE)

-0.032

-0.033

-0.213**

-0.255***

~

(0.018)

(0.018)

(0.073)

(0.077)

Merit: Effort (BE)

0.177***

0.173***

-0.062

0.179***

~

(0.032)

(0.031)

(0.097)

(0.032)

Merit: Talent (BE)

0.015

0.021

0.014

-0.266**

~

(0.031)

(0.031)

(0.031)

(0.099)

Merit: Effort (WE) x Perception inequality (WE)

0.017

~

~

~

~

(0.012)

~

~

~

Merit: Talent (WE) x Perception inequality (WE)

~

0.011

~

~

~

~

(0.012)

~

~

Merit: Effort (BE) x Perception inequality (BE)

~

~

0.072**

~

~

~

~

(0.028)

~

Merit: Talent (BE) x Perception inequality (BE)

~

~

~

0.082**

~

~

~

~

(0.027)

Controls

Yes

Yes

Yes

Yes

BIC

20842.694

20835.920

20808.677

20806.384

Numb. obs.

8643

8643

8643

8643

Num. groups: individuals

1687

1687

1687

1687

Var: individuals (Intercept)

0.145

0.145

0.141

0.141

Var: individuals, merit effort cwc

0.020

~

~

~

Var: individuals, perception inequality cwc

0.000

0.000

~

~

Cov: individuals (Intercept), merit effort cwc

0.004

~

~

~

Cov: individuals (Intercept), perception inequality cwc

-0.006

-0.005

~

~

Cov: individuals, merit effort cwc, perception inequality cwc

0.000

~

~

~

Var: Residuals

0.512

0.511

0.528

0.528

Var: individuals, merit talent cwc

~

0.020

~

~

Cov: individuals (Intercept), merit talent cwc

~

0.012

~

~

Cov: individuals, merit talent cwc, perception inequality cwc

~

0.001

~

~

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05. CWC = centered within group.

@@

~

Model 12

Model 13

Model 14

Model 15

Model 16

Model 17

Intercept

1.543***

1.527***

1.521***

1.680***

1.374***

1.362***

~

(0.115)

(0.114)

(0.114)

(0.139)

(0.133)

(0.134)

Wave x Perception inequality (WE)

-0.004

~

~

~

~

~

~

(0.006)

~

~

~

~

~

Wave x Merit: effort (WE)

~

-0.008

~

~

~

~

~

~

(0.006)

~

~

~

~

Wave x Merit: talent (WE)

~

~

-0.006

~

~

~

~

~

~

(0.006)

~

~

~

Wave x Perception inequality (BE)

~

~

~

0.011

~

~

~

~

~

~

(0.007)

~

~

Wave x Merit: effort (BE)

~

~

~

~

-0.021*

~

~

~

~

~

~

(0.008)

~

Wave x Merit: talent (BE)

~

~

~

~

~

-0.020*

~

~

~

~

~

~

(0.008)

Controls

Yes

Yes

Yes

Yes

Yes

Yes

BIC

20792.207

20767.371

20754.324

20782.059

20749.596

20754.417

Numb. obs.

8643

8643

8643

8643

8643

8643

Num. groups: individuals

1687

1687

1687

1687

1687

1687

Var: individuals (Intercept)

0.176

0.167

0.168

0.383

0.254

0.010

Var: individuals, perception inequality cwc

0.000

~

~

~

~

~

Var: individuals, wave

0.010

0.009

0.009

0.009

0.009

0.009

Cov: individuals (Intercept), perception inequality cwc

-0.007

~

~

~

~

~

Cov: individuals (Intercept), wave

-0.021

-0.018

-0.018

-0.048

-0.003

0.006

Cov: individuals, perception inequality cwc, wave

0.002

~

~

~

~

~

Var: Residuals

0.495

0.480

0.479

0.494

0.494

0.494

Var: individuals, merit effort cwc

~

0.019

~

~

~

~

Cov: individuals (Intercept), merit effort cwc

~

0.023

~

~

~

~

Cov: individuals, merit effort cwc, wave

~

-0.005

~

~

~

~

Var: individuals, merit talent cwc

~

~

0.019

~

~

~

Cov: individuals (Intercept), merit talent cwc

~

~

0.034

~

~

~

Cov: individuals, merit talent cwc, wave

~

~

-0.006

~

~

~

Var: individuals, perception inequality mean

~

~

~

0.008

~

~

Cov: individuals (Intercept), perception inequality mean

~

~

~

-0.045

~

~

Cov: individuals, perception inequality mean, wave

~

~

~

0.009

~

~

Var: individuals, merit effort mean

~

~

~

~

0.053

~

Cov: individuals (Intercept), merit effort mean

~

~

~

~

-0.089

~

Cov: individuals, merit effort mean, wave

~

~

~

~

-0.006

~

Var: individuals, merit talent mean

~

~

~

~

~

0.025

Cov: individuals (Intercept), merit talent mean

~

~

~

~

~

-0.007

Cov: individuals, merit talent mean, wave

~

~

~

~

~

-0.009

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05. CWC = centered within group.

@@

@@

>>>>

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra-Sanhueza}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility.

@@

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature
(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
\citeproc{ref-pierson_when_1993}{Pierson, 1993}).

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they endorse
meritocratic principles. This distinction aligns with recent findings
indicating that individuals distinguish between how merit is perceived
in society and how it should ideally operate, which in turn shapes their
preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}). Meritocratic beliefs serve as symbolic justifications for
unequal outcomes, particularly when access is stratified by income or
social opportunity. Prior studies have shown that individuals who
perceive higher levels of meritocracy tend to express stronger support
for unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

@@

\caption{}
\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
\hline
Intercept                          & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
                                   & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
Wave (Ref.= 2016)                  &                &                &                &                &                &                &                \\
                                   &                &                &                &                &                &                &                \\
\quad Wave 2017                    & $-0.167^{***}$ &                &                &                &                &                &                \\
                                   & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2018                    & $-0.025$       &                &                &                &                &                &                \\
                                   & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2019                    & $-0.046$       &                &                &                &                &                &                \\
                                   & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2022                    & $0.285^{***}$  &                &                &                &                &                &                \\
                                   & $(0.028)$      &                &                &                &                &                &                \\
\quad Wave 2023                    & $0.284^{***}$  &                &                &                &                &                &                \\
                                   & $(0.027)$      &                &                &                &                &                &                \\
Wave                               &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
                                   &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
Wave$^2$                           &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
                                   &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
Perception inequality (WE)         &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
                                   &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
Merit: Effort (WE)                 &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Merit: Talent (WE)                 &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Perception inequality (BE)         &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
                                   &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
Merit: Effort (BE)                 &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
Merit: Talent (BE)                 &                &                &                &                &                & $0.040$        & $0.019$        \\
                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
\hline
Controls                           & No             & No             & No             & No             & No             & No             & Yes            \\
BIC                                & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
Numb. obs.                         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
Num. groups: individuals           & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
Var: individuals (Intercept)       & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
Var: Residual                      & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
Var: individuals, wave             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
Cov: individuals (Intercept), wave & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
\hline
\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
\end{tabular}
\label{table:coefficients}
\end{center}

@@

\caption{}
\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c c c}
\hline
 & Model 8 & Model 9 & Model 10 & Model 11 \\
\hline
Intercept                                                     & $1.561^{***}$  & $1.555^{***}$  & $2.187^{***}$  & $2.340^{***}$  \\
                                                              & $(0.115)$      & $(0.115)$      & $(0.265)$      & $(0.281)$      \\
Perception inequality (WE)                                    & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ \\
                                                              & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
Merit: Effort (WE)                                            & $0.072^{***}$  & $0.074^{***}$  & $0.073^{***}$  & $0.073^{***}$  \\
                                                              & $(0.013)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Merit: Talent (WE)                                            & $0.031^{*}$    & $0.031^{*}$    & $0.030^{*}$    & $0.030^{*}$    \\
                                                              & $(0.012)$      & $(0.013)$      & $(0.012)$      & $(0.012)$      \\
Perception inequality (BE)                                    & $-0.032$       & $-0.033$       & $-0.213^{**}$  & $-0.255^{***}$ \\
                                                              & $(0.018)$      & $(0.018)$      & $(0.073)$      & $(0.077)$      \\
Merit: Effort (BE)                                            & $0.177^{***}$  & $0.173^{***}$  & $-0.062$       & $0.179^{***}$  \\
                                                              & $(0.032)$      & $(0.031)$      & $(0.097)$      & $(0.032)$      \\
Merit: Talent (BE)                                            & $0.015$        & $0.021$        & $0.014$        & $-0.266^{**}$  \\
                                                              & $(0.031)$      & $(0.031)$      & $(0.031)$      & $(0.099)$      \\
Merit: Effort (WE) x Perception inequality (WE)               & $0.017$        &                &                &                \\
                                                              & $(0.012)$      &                &                &                \\
Merit: Talent (WE) x Perception inequality (WE)               &                & $0.011$        &                &                \\
                                                              &                & $(0.012)$      &                &                \\
Merit: Effort (BE) x Perception inequality (BE)               &                &                & $0.072^{**}$   &                \\
                                                              &                &                & $(0.028)$      &                \\
Merit: Talent (BE) x Perception inequality (BE)               &                &                &                & $0.082^{**}$   \\
                                                              &                &                &                & $(0.027)$      \\
\hline
Controls                                                      & Yes            & Yes            & Yes            & Yes            \\
BIC                                                           & $20842.694$    & $20835.920$    & $20808.677$    & $20806.384$    \\
Numb. obs.                                                    & $8643$         & $8643$         & $8643$         & $8643$         \\
Num. groups: individuals                                      & $1687$         & $1687$         & $1687$         & $1687$         \\
Var: individuals (Intercept)                                  & $0.145$        & $0.145$        & $0.141$        & $0.141$        \\
Var: individuals, merit effort cwc                            & $0.020$        & $$             & $$             & $$             \\
Var: individuals, perception inequality cwc                   & $0.000$        & $0.000$        & $$             & $$             \\
Cov: individuals (Intercept), merit effort cwc                & $0.004$        & $$             & $$             & $$             \\
Cov: individuals (Intercept), perception inequality cwc       & $-0.006$       & $-0.005$       & $$             & $$             \\
Cov: individuals, merit effort cwc, perception inequality cwc & $0.000$        & $$             & $$             & $$             \\
Var: Residuals                                                & $0.512$        & $0.511$        & $0.528$        & $0.528$        \\
Var: individuals, merit talent cwc                            & $$             & $0.020$        & $$             & $$             \\
Cov: individuals (Intercept), merit talent cwc                & $$             & $0.012$        & $$             & $$             \\
Cov: individuals, merit talent cwc, perception inequality cwc & $$             & $0.001$        & $$             & $$             \\
\hline
\multicolumn{5}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
\end{tabular}
\label{table:coefficients}
\end{center}

@@

\caption{}
\begin{center}
\scalebox{0.7}{
\begin{tabular}{l c c c c c c}
\hline
 & Model 12 & Model 13 & Model 14 & Model 15 & Model 16 & Model 17 \\
\hline
Intercept                                                & $1.543^{***}$ & $1.527^{***}$ & $1.521^{***}$ & $1.680^{***}$ & $1.374^{***}$ & $1.362^{***}$ \\
                                                         & $(0.115)$     & $(0.114)$     & $(0.114)$     & $(0.139)$     & $(0.133)$     & $(0.134)$     \\
Wave x Perception inequality (WE)                        & $-0.004$      &               &               &               &               &               \\
                                                         & $(0.006)$     &               &               &               &               &               \\
Wave x Merit: effort (WE)                                &               & $-0.008$      &               &               &               &               \\
                                                         &               & $(0.006)$     &               &               &               &               \\
Wave x Merit: talent (WE)                                &               &               & $-0.006$      &               &               &               \\
                                                         &               &               & $(0.006)$     &               &               &               \\
Wave x Perception inequality (BE)                        &               &               &               & $0.011$       &               &               \\
                                                         &               &               &               & $(0.007)$     &               &               \\
Wave x Merit: effort (BE)                                &               &               &               &               & $-0.021^{*}$  &               \\
                                                         &               &               &               &               & $(0.008)$     &               \\
Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.020^{*}$  \\
                                                         &               &               &               &               &               & $(0.008)$     \\
\hline
Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
BIC                                                      & $20792.207$   & $20767.371$   & $20754.324$   & $20782.059$   & $20749.596$   & $20754.417$   \\
Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
Var: individuals (Intercept)                             & $0.176$       & $0.167$       & $0.168$       & $0.383$       & $0.254$       & $0.010$       \\
Var: individuals, perception inequality cwc              & $0.000$       & $$            & $$            & $$            & $$            & $$            \\
Var: individuals, wave                                   & $0.010$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       \\
Cov: individuals (Intercept), perception inequality cwc  & $-0.007$      & $$            & $$            & $$            & $$            & $$            \\
Cov: individuals (Intercept), wave                       & $-0.021$      & $-0.018$      & $-0.018$      & $-0.048$      & $-0.003$      & $0.006$       \\
Cov: individuals, perception inequality cwc, wave        & $0.002$       & $$            & $$            & $$            & $$            & $$            \\
Var: Residuals                                           & $0.495$       & $0.480$       & $0.479$       & $0.494$       & $0.494$       & $0.494$       \\
Var: individuals, merit effort cwc                       & $$            & $0.019$       & $$            & $$            & $$            & $$            \\
Cov: individuals (Intercept), merit effort cwc           & $$            & $0.023$       & $$            & $$            & $$            & $$            \\
Cov: individuals, merit effort cwc, wave                 & $$            & $-0.005$      & $$            & $$            & $$            & $$            \\
Var: individuals, merit talent cwc                       & $$            & $$            & $0.019$       & $$            & $$            & $$            \\
Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.034$       & $$            & $$            & $$            \\
Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.006$      & $$            & $$            & $$            \\
Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.008$       & $$            & $$            \\
Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $-0.045$      & $$            & $$            \\
Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $0.009$       & $$            & $$            \\
Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.053$       & $$            \\
Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.089$      & $$            \\
Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.006$      & $$            \\
Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.025$       \\
Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.007$      \\
Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.009$      \\
\hline
\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
\end{tabular}
\label{table:coefficients}
\end{center}

@@

\bibitem[\citeproctext]{ref-campbell_policy_2012}
Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
\url{https://doi.org/10.1146/annurev-polisci-012610-135202}

@@

\bibitem[\citeproctext]{ref-pierson_when_1993}
Pierson, P. (1993). When {Effect Becomes Cause}: {Policy Feedback} and
{Political Change}. \emph{World Politics}, \emph{45}(4), 595--628.
\url{https://doi.org/10.2307/2950710}
